---
title: "Homework 4"
author: "Your name here"
date: 'Assigned: September 24, 2015'
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 5
    fig_height: 5
---

### Homework outline

This homework is designed to give you practice with writing a more complex function.  In particular, your final function will make calls to other simpler functions that you write along the way.  The overall task has been broken up into sub-tasks to help you decompose the problem into smaller, more tractable components.

> **Goal**: You will produce a function that generates what's often called 'Table 1', which is intended to provide summary information about the variables in your data, broken down by some factor of interest.  

Problem 5 will have you apply the function to the `titanic3` data, which is available in the `PASWR` library.

```{r}
library(PASWR)
str(titanic3)
```

You'll use your function on the `titanic3` data to produce the following table, which summarizes information about passengers on the Titanic, displaying summaries separately depending on the passenger's fare class (1st, 2nd or 3rd).

|Variable                  |Missing |1st           |2nd           |3rd           |P-value |
|:-------------------------|:-------|:-------------|:-------------|:-------------|:-------|
|Survival rate             |0       |61.9%         |43%           |25.5%         |0       |
|% Female                  |0       |44.6%         |38.3%         |30.5%         |0       |
|Age                       |263     |39.16 (14.55) |29.51 (13.64) |24.82 (11.96) |0       |
|# siblings/spouses aboard |0       |0.44 (0.61)   |0.39 (0.59)   |0.57 (1.3)    |0.0279  |
|# children/parents aboard |0       |0.37 (0.72)   |0.37 (0.69)   |0.4 (0.98)    |0.779   |
|Fare ($)                  |1       |87.51 (80.45) |21.18 (13.61) |13.3 (11.49)  |0       |

From this table we can conclude things such as:

  - The Fare was unknown for just 1 of the passengers
  - There is a statistically significant difference in the proportion of female passengers across the fare classes (p-value < 0.0001).  1st class had 44.6% female passengers, compared to 30.5% in 3rd class.

### Important notes

#### Handling missing values

You will be encountering missing values in this homework (entries coded `NA`).  To make sure that not all of your calculations turn out to be `NA`, you'll need to be careful about how you treat missing values.

Example:  When using the `mean()` function, you'll want to use `mean(..., na.rm=TRUE)`, or some alternative.  

```{r}
x <- c(NA, 1, 5, 3, NA, 2)
mean(x)
mean(x, na.rm=TRUE) # Remove missing values
mean(x[!is.na(x)]) # Select only non-missing values
```

Functions like `t.test()` have an `na.action` argument.  The default is to use whatever is returned by `getOption("na.action")`.  You'll want this to be `"na.omit"`, which is the default.

#### Testing your functions

Just as with Homework 3, I will be providing you with a list of test cases on which to test your function.  The test cases and printouts of expected output are available in a supplementary document on the course website.

### Problem 1: Summarizing numeric data

Write a function with the following specification:

1. Input: 2 inputs
    - `x`: numeric vector
    - `groups`: factor vector of the same length as `x`

2. Output: a list with the following elements:
    - `missing`: the number of values of `x` that are missing
    - `means`: means of `x` across the different levels of `groups`
    - `sds`: standard deviations of `x` across the different levels of `groups`
    - `p.value`: p-value output by a test of whether the mean of `x` is the same across all levels of `groups`  (i.e., p-value from a t-test or 1-way ANOVA)
    - `is.binary = FALSE`: a field called `is.binary`, which is set to `FALSE`
    
Here is an example of the output (in this example, the input had 2 missing values, and `groups` had 3 levels)

```
$missing
[1] 2

$means
[1]  0.06646450 -0.01969902  1.96278957

$sds
[1] 0.5800607 1.4293022 0.7123807

$p.value
[1] 0.01107936

$is.binary
[1] FALSE
```

```{r}
library(plyr)

generatepvalueforNumeric <- function(x, groups){
   if(nlevels(groups) > 2){
            p.value.result <- summary(aov(x ~ groups))[[1]][["Pr(>F)"]][1]
        }
    else if(nlevels(groups) <= 2){
            p.value.result <- t.test(x ~ groups, na.action = "na.omit")$p.value
        }
      return(p.value.result)
  }

generateNumericSummary <- function(x, groups) {
  missing <- length(x[is.na(x)])
  df <- data.frame(x, groups)

  group.vars <- ddply(df, ~groups, summarize, means = mean(x, na.rm = TRUE), sds = sd(x,na.rm = TRUE))
  
  result.list <- list(missing = missing, means = group.vars$means, sds = group.vars$sds, p.value = generatepvalueforNumeric(x, groups), is.binary = FALSE)
  return (result.list)
}
```

**Hint:** Missing values in R are coded as `NA`.  You can use the function `is.na(x)` to test if `x` is missing.

### Problem 2: Summarizing binary data

Write a function with the following specification:

1. Input: 2 inputs
    - `x`: binary (0/1) vector
    - `groups`: factor vector of the same length as `x`

2. Output: a list with the following elements:
    - `missing`: the number of values of `x` that are missing
    - `prop`: the percent of non-NA `x` values that equal 1 in each level of `groups`
    - `p.value`: p-value output by a test of whether the proportion of `x` that is equal to 1 is the same across all levels of `groups`  (i.e., p-value from a fisher.test)
    - `is.binary = TRUE`: a field called `is.binary`, which is set to `TRUE`

Here is an example of the output (in this example, the input had 4 missing values, and `groups` had 3 levels)
```
$missing
[1] 4

$prop
[1] 0.3333333 0.0000000 0.6000000

$p.value
[1] 0.2525253

$is.binary
[1] TRUE
```

```{r}
generatepvalueforBinary <- function(x, groups){
  p.value.result <- fisher.test(x, groups)$p.value
}

prop <- function(x){
  return (mean(x, na.rm = TRUE))
}

generateBinarySummary <- function(x, groups) {
  missing <- length(x[is.na(x)])
  df <- data.frame(x, groups)
  
  group.vars <- ddply(df, ~groups, summarize, prop = prop(x))
  result.list <- list(missing = missing, prop = group.vars$prop, p.value = generatepvalueforBinary(x, groups), is.binary = TRUE)
  return (result.list)
}
```

### Problem 3: Getting the right data type

Write a function with the following specification:

1. Input: 2 inputs
    - `x`: a vector
    - `groups`: factor vector of same length as `x`

2. Output:
    - If `x` is binary, return `generateBinarySummary(x, groups)`
    - Else, if `x` is numeric, return `generateNumericSummary(x, groups)`
    - Else, return `NULL`
    
Here's an example of output when `x` is numeric and `groups` has 2 levels
```
$missing
[1] 2

$means
[1] 0.02816961 1.96278957

$sds
[1] 0.9676708 0.7123807

$p.value
[1] 0.001395812

$is.binary
[1] FALSE
```

Here's an example of output when `x` is binary and `groups` has 2 levels

```
$missing
[1] 4

$prop
[1] 0.1428571 0.6000000

$p.value
[1] 0.2222222

$is.binary
[1] TRUE
```
    
```{r}
  isBinary <- function(x){
    x <- x[which(!is.na(x))]
    return (all(x == 0 | x == 1))
  }

  generateVariableSummary <- function(x, groups) {
    if(isBinary(x)){
      generateBinarySummary(x, groups)
    }
    else if(is.numeric(x)) {
      generateNumericSummary(x, groups)
    }
    else 
      NULL
  }
```

### Problem 4: Formatting output
Write a function with the following specification:

1. Input: 1 input
   - `var.summary`: the output of a `generateVariableSummary` call

2. Output: 
   - If `var.summary` is for a numeric variable and the grouping variable has `k` levels, return a vector of the form
   (`num.missing`, `"mean[1] (sd[1])"`, `"mean[2] (sd[2])"`, ..., `"mean[k] (sd[k])"`, `p.value`)
   
   - If `var.summary` is for a binary variable and the grouping variable has `k` levels, return a vector of the form
   (`num.missing`, `"prop[1]%"`, `"prop[2]%"`, ..., `"prop[k]%"`, `p.value`)
   
Here's sample output when `x` is numeric and `groups` has 3 levels

```
[1] "0"  "0.23 (0.91)" "1.3 (0.81)"  "0.93 (1.2)"  "0.2218"  
```
Here's sample output when `x` is binary and `groups` has 2 levels

```
"4"   "14.3%"  "60%"    "0.2222"
```

```{r}

formatVariableSummary <- function(var.summary) {
   if(!is.null(var.summary))
   {  
      result.summary <- c(var.summary$missing)
      if(var.summary$is.binary){
        for(i in 1:length(var.summary$prop)){
          prop.comp <- paste(round(var.summary$prop[i]*100, 1), "%", sep="")
          result.summary <- paste(result.summary,prop.comp  ,sep = " ")
        }
      }
      else {
        for(i in 1:length(var.summary$means)){
          ith.sd <- paste("(", round(var.summary$sds[i],2), ")", sep = "")
          ith.mean.and.sd <- paste(round(var.summary$means[i],1), ith.sd, sep = "")
          result.summary <- paste( result.summary, ith.mean.and.sd, sep = " ")
        }
      }
      result.summary <- paste(result.summary, round(var.summary$p.value, 4), sep=" ")
      return (strsplit(result.summary, " ")[[1]])
  }
  else return (NULL)
}
```

**Hints:**  Given two numbers `a` and `b`, here's how you get the formatting for `"mean[1] (sd[1])"`.  This gives an example where the mean is 21.89 and the sd is 1.83.

```{r}
a <- 21.89
b <- 1.83
paste(a, " (", b , ")", sep="")
```

Similarly, given a percentage `a`, here's how to get R to return `"a%"`:

```{r}
a <- 75.2
paste(a, "%", sep="")
```

Alternatively, if you're familiar with `sprintf()` from C or other languages, you are free to use that.

### Problem 5: Piecing things together

**(a)** Using the functions you write in Problems 1 - 4, write a function with the following specification:

1. Input: 3 inputs
    - `dataset`: a data frame
    - `var.names`: a character vector that is a subset of `colnames(dataset)`
    - `group.name`: a string that gives the name of the grouping variable
   
2. Output:
    - Your function should return a data frame that has the following form (this is an example where we have three variables, one numeric and two binary, and a **group variable with 2 levels**):
    
|  Variable    | Missing | groups[1] | groups[2]  | P-value |
|:------------:|:-------:|:---------:|:----------:|:-------:|
| numeric.var  |   12    |45.2 (2.7) | 53.9 (4.1) |  0.0142 |
| binary.var1  |    0    |  14%      |   21%      |  0.0679 |
| binary.var2  |    7    |  95%      |   94%      |  0.7810  |

  - The data frame will have 3 + #(levels in grouping variable) columns
  - The first column will give the variable names specified in `var.names`
  - The second column will give the number of missing values
  - The next #(levels in grouping variable) columns will give the group means and standard errors (as shown) if the variable is **numeric**, or % of 1's if the variable is **binary**.
  - The last column will give the p-value for testing whether the group means (or proportions) are the same across all levels of the grouping variable.
  
  
```{r}
generateDataSummary <- function(dataset, varnames, group.name) {
  result <- data.frame()
  for(i in 1:length(varnames)){
     output <- formatVariableSummary(generateVariableSummary(dataset[[varnames[i]]], dataset[[group.name]]))
    if(!is.null(output)){
      sub.result <- data.frame("Variable" = varnames[i], "Missing" = output[1])
      for(j in 1:length(levels(dataset[[group.name]]))) {
        sub.result[levels(dataset[[group.name]])[j]] <- output[1 + j] 
      }
      sub.result["P-value"] <-  output[(length(output))]
      result <- rbind(result, sub.result)
     }
  }
  kable(return (result), format = "markdown")
}
```
**Hint**: One way of doing this is to use a loop and the `rbind()` function to build up your summary table row by row.

**(b)** Using the `kable` function, create a nice summary table obtained by running `generateDataSummary()` on the `titanic4` data defined below, which is a transformation of the `titanic3` data in the `PASWR` library.  I've gotten you started by pre-processing the data a little bit.  


```{r}
# Load libraries
library(PASWR)
library(plyr)
library(knitr)

# Construct a variable called 'female' which is a binary variable indicating
# whether the passenger is female.
# Store this modified data frame as titanic4
titanic4 <- transform(titanic3, female = as.numeric(sex == "female"))

# Use the following var names to form your table:
var.names <- c("survived", "female", "age", "sibsp", "parch", "fare")

# Use passenger glass (pclass) as your grouping variable.
group.name <- "pclass"
```

```{r, results = 'asis'}
kable(generateDataSummary(titanic4, var.names, group.name)  , format = "markdown")
```


**(c)** Look at the survival rate row of your table and answer the following questions.

- What passenger class has the highest survival rate?  
     - Passenger who was in the first class

- What is the survival rate for this passenger class?  
     - 61%
     
- Are survival rates statistically significantly different across the passenger classes?  
     - Yes

- What statistical test was used to obtain the p-value?
     - Fisher

**(d)** Look at the age row of your table and answer the following questions.  

- For how many passengers in the data is the age unknown?  
     - 263

- Is there a statistically significant difference in average passenger age across the passenger classes?  
     - Yes

- What statistical test was used to obtain the p-value?
     - T-Test
