---
title: "Homework 1"
author: "Anisha Kaul"
date: 'Assigned: September 3, 2015'
output: html_document
---

##### This homework is due by **1:30pm on Thursday, September 10**.  To complete this assignment, follow these steps:

1. Download the `homework1.Rmd` file from Blackboard or the course website.

2. Open `homework1.Rmd` in RStudio.

3. Replace the "Your Name Here" text in the `author:` field with your own name.

4. Supply your solutions to the homework by editing `homework1.Rmd`.

5. When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `homework1_YourNameHere.Rmd`, and submit on Blackboard.  (YourNameHere should be changed to your own name.)

##### Homework tips:

1. Recall the following useful RStudio hotkeys.

Keystroke | Description
----------|-------------
`<tab>` | Autocompletes commands and filenames, and lists arguments for functions.
`<up>` | Cycles through previous commands in the console prompt
`<ctrl-up>` | Lists history of previous commands matching an unfinished one
`<ctrl-enter>` | Runs current line from source window to Console. Good for trying things out ideas from a source file.
`<ESC>` | Aborts an unfinished command and get out of the + prompt

**Note**: Shown above are the Windows/Linux keys.  For Mac OS X, the `<ctrl>` key should be substituted with the `<command>` (&#8984;) key.

2. Instead of sending code line-by-line with `<ctrl-enter>`, you can send entire code chunks, and even run all of the code chunks in your .Rmd file. Look under the <Chunks> menu of the Source panel.

3. Run your code in the Console and Knit HTML frequently to check for errors.

4. You may find it easier to solve a problem by interacting only with the Console at first. 

### Homework 1 outline

This homework gets you to create a "Cheat Sheet" that you can refer back to over the course of the semester. 

### Problem 1: Simple Boolean operations

> Tip:  Note that each of the code blocks in this Problem contain the expression `eval = FALSE`.  This tells R Markdown to display the code contained in the block, but not to evaluate it.  To check that your answer makes sense, be sure to try it out in the console with various choices of values for the variable `x`.   


##### (a) Checking equality.

Given a variable `x`, write a Boolean expression that returns `TRUE` if the variable `x` is equal to "cat".  

```{r, eval = FALSE}
x == "cat"
```

##### (b) Checking inequality.

Given a variable `x`, write a Boolean expression that returns `TRUE` if the variable `x` is *not* `NA` (i.e., is not missing).

```{r, eval = FALSE}
is.na(x)
```

##### (c) Checking if a number is in a given range. 

Given a (possibly negative) number `x`, write a Boolean expression that returns `TRUE` if and only if `x` is smaller than `-40` or bigger than `105`.  

```{r, eval=FALSE}
(x < -40 ) | (x > 105)
```

##### (d) A more complicated example.

Given an integer number `x`, write a Boolean expression that returns `TRUE` if and only if `x` is an **even** number between 17 and 25 or 77 and 101.  

```{r, eval=FALSE}
(x%%2 == 0) & ( (x >= 17 & x <= 25) | (x >= 77 & x <= 101))
```

**Tip:** Recall the modulus operator we saw in lecture 1: `%%`.  For integers `x` and `y`, `x %% y` is the remainder of `x` divided by `y`.  

### Problem 2: Vector Boolean operations

##### (a) R has two kinds of Boolean operators implemented, single (`&`, `|`) and double (`&&`, `||`).  

One of these operators takes advantage of something called *lazy evaluation* while the other does not.  They also don't behave the same way when applied to *vectors*.  

Read the help file (`help("||")`) and construct some examples to help figure out how the two behave.  

To help you get started, try out the following two examples in your console:

```{r, eval = FALSE}
# Example:  The variable y.prob2a is never defined.  
# (Do not define it!)
# What happens when you run this code?
x.prob2a <- 5
(x.prob2a < 10) | (y.prob2a > 2)
(x.prob2a < 10) || (y.prob2a > 2)
```

```{r, eval = FALSE}
# Define vectors
x.prob2a.vec <- c(TRUE, FALSE, FALSE)
y.prob2a.vec <- c(TRUE, TRUE, FALSE)

# Apply various Boolean operations to see what happens
x.prob2a.vec & y.prob2a.vec
x.prob2a.vec && y.prob2a.vec
x.prob2a.vec | y.prob2a.vec
x.prob2a.vec || y.prob2a.vec
```

Can you explain what's happening?  Write up a brief explanation below.

The && or the || operator performs the logical operation on only the first element of the vectors(compares only the first element of each vector) -> lazy evaluation

& or the | evaluates and compares all the components. For this expression to evaluate, both the vectors have to have the same length



##### (b) Using `any()`

Two people were asked to give their preferences between two options: [Facebook, Twitter], [Firefox, Chrome], [Mac, PC], [Summer, Winter].  Their results are given below.

```{r}
choices1 <- c("Twitter", "Chrome", "Mac", "Summer")
choices2 <- c("Facebook", "Chrome", "PC", "Summer")
```

Use the `any()` function to determine if the two people have any preferences in common.  (Your code should output a single Boolean value, either `TRUE` or `FALSE`)
```{r}
any(choices1 == choices2)
```

##### (c) Using `all()` 

Use the `all()` function to determine if the two people have identical preferences. (Your code should ouput a single Boolean value, either `TRUE` or `FALSE`)

```{r}
all(choices1 == choices2)
```

##### (d) Missing values.

Let `age` be the vector defined below.

```{r}
age <- c(37, 21, 92, NA, 45, NA, NA, 18)
```

Write a Boolean expression that checks whether each entry of `age` is missing (recall missing values are denoted by `NA`).  Your expression should return a Boolean vector having the same length as `age`.

```{r}
is.na(age)
```

### Problem 3: Referencing vector elements

##### (a)  `which()` practice

Write code that returns the indexes of `age` that are missing.

```{r}
which(is.na(age))
```

##### (b) Getting non-missing values

Write code that uses negative indexes and your solution from (a) to return only the values of `age` that are *not* missing. (i.e., your code should result in a vector with elements: 37, 21, 92, 45, 18) 


```{r}
age[- c(which((is.na(age))))]
```

##### (c)  A more direct way of getting non-missing values

Using the negation operator `!` and the `is.na()` function, write an expression that returns only the values of `age` that are *not* missing.

```{r}
age[which(!is.na(age))]
```

##### (d) More `which()` practice

For the next three problem we'll go back to the `cars` data set.

```{r}
speed <- cars$speed
dist <- cars$dist
```

Write code to figure out which cars had a stopping distance of 75 feet or more.

```{r}
cars[which(cars["dist"] >= 75), ]
```

##### (e) `which.min`, `which.max` practice

Use the `which.min()` function to figure out which car had the *shortest* stopping distance. (Your code should return the car's index.)

```{r}
which.min(cars$dist)
```

##### (f) More practice 

Use the `which.max()` function to figure out the *speed* of the car that had the *longest* stopping distance.  (Your code should return the car's speed.)

```{r}
which.max(cars$speed)
```

### Problem 4: Data frame basics

##### (a) Importing data.

Use the `read.csv()` function to import the survey data into a variable called `survey`.

```{r}
survey <- read.csv("http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data.csv")
```

##### (b) `$` notation

Use the `$` operator to select the TVhours column from the `survey` data

```{r}
survey$TVhours
```

##### (c) [,] notation

Repeat part (b) using `[,]` notation.  i.e., Use `[,]` notation to select the TVhours column from the `survey` data by name (i.e., obtain this column by using the name "TVhours" instead of using the column number)

```{r}
survey[,"TVhours"]
```

##### (d) [[]] notation

Repeat part (c) with [[]] notation.

```{r}
survey[["TVhours"]]
```

##### (e) [] notation

Repeat part (d), but this time using single blackets (`[ ]`) notation.  

(Observe that this returns a new single-column *data frame*, not just a vector.) 

```{r}
survey["TVhours"]
```

##### (f) `subset()` practice

Use the `subset()` function to select all the survey data on Program and OperatingSystem for respondents whose Rexperience is "Never used" *or* who watched 5 or more hours of TV last week.

```{r}
subset(survey, Rexperience == "Never used" | TVhours >= 5, select = c("Program", "OperatingSystem"))
```


### Problem 5: Data summaries and inline code practice.

##### (a) Bar graph

Create a bar graph of respondents' Rexperience.

```{r, fig.align='center', fig.width=7, fig.height=4}
plot(survey["Rexperience"])
```

##### (b) Inline code practice

Replace all occurrences of ???? in the paragraph below with an inline code chunk supplying the appropriate information.

> Of the  `r dim(survey)[1]` survey respondents, `r dim(subset(survey, Program == "MISM"))[1]` were from the MISM program.  We found that `r round((dim(subset(survey, OperatingSystem == "Mac OS X"))[1] / dim(survey)[1] ) *100 , 1)`% of the all students in the class use the Mac OS X operating system.
